Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-06-17 analysis/paper_draft4.Rmd
added references to NJ, CT, Batson ultimate-inquiry.

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  United States law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal strike bias by showing how to incorporate historical data on an attorney's use of peremptory strikes in past cases. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences bias detection (how likely the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the United States Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or membership in another category that the law prohibits as a basis for striking that juror. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)]. Such illegal strike bias is the parameter of interest, not to be confused with "bias" in a statistical sense, or bias in a parameter [EDITME].

@@

For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes (Wash. General Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1); Conn. Superior Court Rule 5-12(d)). Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@Williams2017].

@@

To be sure, neither this nor any other statistical approach can automatically detect _Batson_ violations, because _Batson_ law permits any evidence of strike bias, not just strikes in past cases. And _Batson_ law  requires the trial judge to weigh all such evidence to decide whether strike bias likely exists (the _Batson_ violation). Still, _before_ accounting for other relevant evidence of strike bias, our approach can help attorneys detect strike bias, given current and historical strike data, a valid model of the relevant court's strike procedure, and a fixed value for the power prior's $\alpha$ parameter. After testing for sensitivity of bias detection to different fixed $\alpha$ values for the power prior [@Carvalho2021, p. 5252; @Ibrahim2015, p. 3734], attorneys can then combine the resulting inference with other evidence to decide whether to argue _Batson_; to prove the _Batson_ prima facie case; and to persuade the trial judge to find  a _Batson_ violation (Figure \@ref(fig:batsonworkflowfig)).

@@

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select for the power prior. Given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

@@

[EDITME]Our simulations demonstrate that when bias is high in either the historical _or_ current data, bias can be detected, especially when $\alpha$ is close to 1. For example, when $b_{curr}$ = 0, $b_{hist}$ = 3, and alpha = 1 our simulation bias detection rate is 1, when there are two or three historical trials (Fig X). In a scenario where we are using real data, and the true values of alpha are not known, these represent cases of extreme incompatibility, and should be detectible through sentivity checking of the value of $\alpha$. For instance, when the value of $\alpha$ is set to 0.1, the bias detection rate goes down to 0.02 for both the two and three historical data set simulations. Such sensitivity to the value of $\alpha$ is indicative of incompatibility between current and historical data.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021, @NJR183A, @CT512
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  United States law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal strike bias by showing how to incorporate historical data on an attorney's use of peremptory strikes in past cases. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences bias detection (how likely the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the United States Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or membership in a cognizable class, i.e., a category that the law prohibits as a basis for striking that juror. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)]. Such illegal strike bias is the parameter of interest, not to be confused with "bias" in a statistical sense, or bias in a parameter [EDITME].

@@

For _Batson_ challenges, the "ultimate inquiry" is whether the striking party "was motivated in substantial part by discriminatory intent." [@flowers2019, p. 2241] In contrast, for similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes (Wash. General Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1); Conn. Superior Court Rule 5-12(d); for a similar rule, see N.J. Court Rule 1:8-3A(d)). Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@Williams2017].

@@

To be sure, neither this nor any other statistical approach can automatically detect _Batson_ violations, because _Batson_ law permits any evidence of strike bias, not just strikes in past cases. And _Batson_ law  requires the trial judge to weigh all such evidence to decide whether illegal strike bias likely exists (the _Batson_ violation). Still, _before_ accounting for other relevant evidence of strike bias, our approach can help attorneys detect strike bias, given current and historical strike data, a valid model of the relevant court's strike procedure, and a fixed value for the power prior's $\alpha$ parameter. After testing for sensitivity of bias detection to different fixed $\alpha$ values for the power prior [@Carvalho2021, p. 5252; @Ibrahim2015, p. 3734], attorneys can then combine the resulting inference with other evidence to decide whether to argue _Batson_; to prove the _Batson_ prima facie case; and to persuade the trial judge to find  a _Batson_ violation (Figure \@ref(fig:batsonworkflowfig)).

@@

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select for the power prior. Given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515) and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

@@

[EDITME]Our simulations demonstrate that when bias is high in either the historical _or_ current data, bias can be detected, especially when $\alpha$ is close to 1. For example, when $b_{curr}$ = 0, $b_{hist}$ = 3, and alpha = 1 our simulation bias detection rate is 1, when there are two or three historical trials (Fig X). In a scenario where we are using real data, and the true values of $\alpha$ are not known, these represent cases of extreme incompatibility, and should be detectible through sensitivity checking of the value of $\alpha$. For instance, when the value of $\alpha$ is set to 0.1, the bias detection rate goes down to 0.02 for both the two and three historical data set simulations. Such sensitivity to the choice of $\alpha$ value indicates incompatibility between current and historical data.
